Estimation of flow trajectories in a multi-lines transportation network

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
update visu of CA
write article

Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally attempt to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems. In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy. The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for overflow; and an update of the marginal flows to avoid transfer overflow. This iterative procedure clearly evokes the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step". The fist step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property). Moreover, this solution offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.

@@

The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example (section \ref{example_construction}) with an additional one. If the estimated pedestrian time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$. The graph structure, consisting in matrices $\bm{\chi}$, $P$ and $T$, was constructed beforehand testing the method.

The algorithm was run with hyperparameter $\theta = 0.1$. Tracking the Mean Margin Error (MME) showed a rapid decrease but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.

@@

\begin{figure}[h]
	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ in geographical clusters.}
	\label{transfers}
\end{figure}

>>>>

Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally attempt to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems. In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy. The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for transfer overflow; and an update of marginal distributions. This iterative procedure clearly evokes the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step". The fist step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property). Moreover, this solution offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.

@@

Line edges were built from the given data, and to obtain the full network structure, we added transfer edges between different lines if the estimated walking time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) was less than $120$ seconds. Shortest-paths between all pairs of stops were computed using the breath-first search algorithm \cite{west2001introduction} implemented in the \emph{igraph}\footnote{\url{https://igraph.org/}} R library \cite{Csardi2006}, thus giving $\bm{\chi}$. The matrix $P$, containing valid $st$-trips, was built using the same conditions found in the toy example experiments (section \ref{example_construction}) with an additional one. If the estimated pedestrian time between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$. The graph structure, consisting in the edge-trip incidence matrix $\bm{\chi}$, the valid $st$-trips set $P$, and the transfer edges set $T$, was constructed beforehand testing the method.

The algorithm was run with hyperparameter $\theta = 0.1$. Tracking the Mean Margin Error (MME) showed a rapid decrease ($<0.001$ after 13 iterations) but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^\text{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.

@@

\begin{figure}[h]
	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ by ``superstop" names.}
	\label{transfers}
\end{figure}
